---
title: "logisticsDataanalysis"
author: "Kanishk Kapoor"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document provides an analysis for a real dataset related to logistics delays and their causes using R.

### 1.Loading Required Packages
```{r}
# Load required package
library(zoo)
library(tidyverse)
library(data.table)
library(readxl)
library(lubridate)
library(ggplot2)
library(corrplot)
library(dplyr)
library(tidyr)
library(tseries)
library(forecast)
library(Metrics)
library(xgboost)
```
### 2. Loading Dataset
```{r}
data <- read.csv("smart_logistics_dataset.csv", stringsAsFactors = FALSE)
head(data)
```
###3. Data Cleaning
```{r}
str(data)

summary(data)

colSums(is.na(data))
```
#### No missing values exist
```{r}
sapply(data,class)
```
#### Conversion to Factors
Attributes such as Asset_ID and timestamp will excluded from conversion to factors
```{r}
# columns to exclude
exclude_cols <- c("Asset_ID", "Timestamp")

#converting rest of the columns as factors
data[, !names(data) %in% exclude_cols] <- lapply(data[, !names(data) %in% exclude_cols], as.factor)
```
```{r}
sapply(data,class)
```
#### Time to POSIXct format using lubridate
```{r}
data$Timestamp <- ymd_hms(data$Timestamp)
data$Timestamp<- as.POSIXct(data$Timestamp,format="%Y-%m-%d %H:%M:%S")

```

### 4.EDA(Exploratory Data Analysis)

#### A.Count Delayed vs non delayed shipments per Truck

```{r}
delay_counts <- data %>%
  group_by(Asset_ID, Logistics_Delay) %>%
  summarise(Count = n(), .groups = "drop") %>%
  
  # Convert to wide format so each truck appears only once
  pivot_wider(names_from = Logistics_Delay, values_from = Count, values_fill = list(Count = 0)) %>%
  
  # Rename columns for clarity
  rename(OnTime_Shipments = `0`, Delayed_Shipments = `1`)

head(delay_counts)
```
```{r}
#Convert from wide to long format for ggplot

long_delay_counts <- delay_counts %>%
  pivot_longer(cols = c(OnTime_Shipments, Delayed_Shipments),
               names_to = "Logistics_Delay",
               values_to = "Count")
#Covert delay labels to 0 and 1 
long_delay_counts$Logistics_Delay <- ifelse(long_delay_counts$Logistics_Delay == "OnTime_Shipments", "0", "1")
```

```{r}
ggplot(long_delay_counts, aes(x= Asset_ID , y = Count, fill= as.factor(Logistics_Delay))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title=" Proportions of Delays Per Truck",
       x = "Truck ID",
       y = "Proportions of Shipments",
       fill = "Logistics Delay(0=NO , 1= YES)") +
    coord_flip()
```

```{r}
#Calculate delay rate per truck
delay_rate<- data%>%
  mutate(Logistics_Delay = as.numeric(as.character(Logistics_Delay))) %>%
  group_by(Asset_ID) %>%
  summarise(Total_Shipments= n(),
            Delayed_Shipments= sum(Logistics_Delay),
            Delay_Rate = Delayed_Shipments/ Total_Shipments) %>%
  arrange(desc(Delay_Rate)) %>%
  head(5)

print(delay_rate)
```
#### MOST DELAYED SHIPMENTS BY TRUCK 10 AND TRUCK 3

#### B. Logistics Delay Reasons
```{r}
delay_reasons<- data %>%
  group_by(Logistics_Delay_Reason) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(delay_reasons)
```
```{r}
weather_delay <- data %>%
  mutate(
    Temperature = as.numeric(as.character(Temperature)),
    Humidity = as.numeric(as.character(Humidity))
  ) %>%
  filter(Logistics_Delay_Reason == "Weather") %>%
  summarise(
    Average_Temperature= mean(Temperature, na.rm= TRUE),
    Average_Humidity = mean(Humidity, na.rm = TRUE)
  )

print(weather_delay)
```
```{r}
ggplot(data, aes(x= Temperature, y=Humidity, color = Logistics_Delay_Reason)) +
  geom_point(alpha=0.5)+
  labs(title="TEMP & HUMIDITY impact on delays",
       x="Temperature",
       y="Humidity")

```

```{r}
ggplot(data, aes(x=Temperature,y=Humidity, color = Logistics_Delay_Reason == "Weather" )) +
  geom_point(alpha = 0.5) +
  labs(title = "Temperature vs Humidity for Weather-related delays",
       x="Temperature",
       y="Humidity") +
  theme_minimal()
```
```{r}
geoData<- data %>%
  filter(Logistics_Delay_Reason != "None") %>%
  mutate(
    Latitude = as.numeric(as.character(Latitude)),
    Longitude = as.numeric(as.character(Longitude))
  )
ggplot(geoData, aes(x = Latitude, y = Longitude)) +
  geom_point(aes(color = Logistics_Delay_Reason), alpha = 0.5) +
  geom_density_2d(aes(color = Logistics_Delay_Reason), linewidth = 0.8, bins=6) +  # Use linewidth instead of size
  labs(title = "Geographical Distribution of Delays",
       x = "Latitude",
       y = "Longitude",
       color = "Delay Reason") +
  theme_minimal()
```
```{r}
ggplot(geoData,aes(x= Latitude, y= Longitude)) +
  geom_bin2d(bins = 50) +
  scale_fill_viridis_c(option = "plasma") +
  geom_point(aes(color = Logistics_Delay_Reason), alpha = 0.5, size =1) +
  labs(title = "Heatmap of logistics delays by location",
       x="Latitude",
       y="Longitude",
       fill="Delay Density",
       color="Delay Reason") +
  theme_minimal()
```

```{r}
traffic_weather <- data %>%
  filter(Logistics_Delay_Reason =="Weather") %>%
  group_by(Traffic_Status) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(traffic_weather)
```

#### C. Seasonal delay patterns
```{r}
# Extract numeric month (1 to 12)
data$Month <- month(data$Timestamp)

# Convert numeric month to abbreviated names
data$Month <- factor(month.abb[data$Month], levels = month.abb)
```

```{r}
monthly_delays <- data %>%
  group_by(Month) %>%
  summarise(Total_Shipments = n(),
            Delayed_Shipments = sum(Logistics_Delay == 1),
            Delay_Rate = Delayed_Shipments / Total_Shipments) %>%
  arrange(match(Month, month.abb))

print(monthly_delays)

```
```{r}
ggplot(monthly_delays, aes(x= Month, y = Delayed_Shipments, fill = Month))+
  geom_bar(stat = "identity") +
  labs(title = "Total Delayed Shipments per Month",
       x="Month",
       y="Number of Delayed Shipments") +
  theme_minimal()
```
##### Delays which were registered because of weather.
```{r}
monthly_weather_delays <- data %>%
  filter(Logistics_Delay_Reason =="Weather") %>%
  group_by(Month) %>%
  summarise(Total_Shipments = n(),
            Delayed_Shipments = sum(Logistics_Delay == 1),
            Delay_Rate = Delayed_Shipments / Total_Shipments) %>%
  arrange(match(Month, month.abb))

print(monthly_weather_delays)
```

```{r}
ggplot(monthly_weather_delays, aes(x= Month, y = Delayed_Shipments, fill = Month))+
  geom_bar(stat = "identity") +
  labs(title = "Total Weather-Delayed Shipments per Month",
       x="Month",
       y="Number of Delayed Shipments") +
  theme_minimal()
```

#### D. Customer Purchase Frequency's correlation with high shipment delays trucks

```{r}
data$User_Purchase_Frequency <- as.numeric(as.character(data$User_Purchase_Frequency))
data$User_Transaction_Amount <- as.numeric(as.character(data$User_Transaction_Amount))

```

```{r}
delayed_trucks <- data %>%
  filter(Asset_ID %in% c("Truck_10","Truck_3")) %>%
  group_by(Month, Asset_ID) %>%
  summarise(
    Avg_Purchase_Frequency = mean(User_Purchase_Frequency, na.rm = TRUE),
    Avg_Transaction_Amount = mean(User_Transaction_Amount, na.rm= TRUE)
  ) %>%
  arrange(Month)

print(delayed_trucks)
    
  
```

##### PURCHASE FREQUENCY OF HIGH DELAY TRUCKS OVER TIME
```{r}
ggplot(delayed_trucks,aes(x=Month, y= Avg_Purchase_Frequency,group = Asset_ID,colour = Asset_ID))+
  geom_line(size=1) +
  geom_point(size=2) +
  labs(title="Impact of Delayed Trucks on user purchase frequency",
       x="Month",
       y="Average Purchase Frequency",
       color="Truck ID") +
  theme_minimal()
      
```
```{r}
ggplot(delayed_trucks,aes(x=Month,y=Avg_Transaction_Amount,group = Asset_ID,color=Asset_ID)) +
  geom_line(size=1) +
  geom_point(size=1) +
  labs(title = "Average Transaction Amount over the year for high delay trucks",
       x= "MONTH",
       y="Average Transaction Amount",
       color="Truck ID") +
  theme_minimal()
  
```

###### AVERAGE USER PUCRHASE FREQUENCY ACROSS THE YEAR FOR ALL THE TRUCKS
```{r}
All_Trucks_Pf <- data %>%
  group_by(Month, Asset_ID) %>%
  summarise(
    Avg_User_Pf = mean(User_Purchase_Frequency, na.rm = TRUE),
    Avg_User_Transaction = mean(User_Transaction_Amount, na.rm= TRUE)
  )

print(All_Trucks_Pf)
```
```{r}
ggplot(All_Trucks_Pf,aes(x=Month,y=Avg_User_Pf, group=Asset_ID,color=Asset_ID)) +
  geom_line(size =1) +
  geom_point(size =1) +
  labs(title = "Impact of All trucks on User Purchase ",
       x= "Month",
       y= "Average Purchase Frequency",
       color ="Truck ID") +
  theme_minimal()
```
```{r}
ggplot(All_Trucks_Pf,aes(x=Month,y=Avg_User_Transaction,group = Asset_ID,color = Asset_ID)) +
  geom_line(size =0.7)+
  geom_point(size=1)+
  labs(title = "Average user transactions over the year of All Trucks",
       x="Month",
       y="Mean Transaction Amount",
       color="Truck ID") +
  theme_minimal()
```
#### E. Asset Utlization Analysis
```{r}
data$Asset_Utilization <-as.numeric(as.character(data$Asset_Utilization))
```

```{r}
asset_util <- data %>%
  group_by(Asset_ID) %>%
  summarise(
    Mean_Asset_Util = mean(Asset_Utilization,na.rm= TRUE),
    Total_Orders = n()
  ) %>%
  arrange(desc(Mean_Asset_Util))

print(asset_util)
  
```
```{r}
ggplot(asset_util,aes(x= reorder(Asset_ID,Mean_Asset_Util), y= Mean_Asset_Util, fill = Mean_Asset_Util)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Mean Asset Utilization per Truck",
       x = "Truck ID",
       y = "Mean Utilization (%)",
       fill = "Utilization %") +
  theme_minimal()
```
```{r}
truck_performance <- delay_rate %>%
  left_join(asset_util, by = "Asset_ID")

print(truck_performance)
```
```{r}
ggplot(truck_performance, aes(x=Mean_Asset_Util, y= Delay_Rate, label= Asset_ID))+
  geom_point(size=3, color="red") +
  geom_text(vjust = -1, size=3.5) +
  labs(title="Asset Utilization vs Delay Rate",
       x="Mean Asset Utilization (%)",
       y="Delay Rate (%)") +
  theme_minimal()
  
```
### 5. Insights & Predictive Modelling
✅ July had the highest shipment delays.(Especially Truck 10 and Truck 3)
✅ User purchase frequency dropped significantly in August & September.
✅ Potential cause-effect relationship: Customers may have lost trust due to delays in July, leading to lower purchase frequencies in the following months.
✅ On the contrary, even though user purchase frequency dropped significantly but 
   the user purchase transaction grew distinctively. 
   POSSIBLE EXPLANATIONS FOR ANOMALY:
1️⃣ Fewer Users, But Bigger Spenders 💰
	•	Regular customers may have stopped purchasing due to delays.
	•	However, a few high-value customers (B2B buyers, bulk orders) may have continued purchasing, increasing total transaction value.

2️⃣ Stockpiling After Delays 📦
	•	Customers affected by July delays might have compensated by making larger purchases later.
	•	Companies ordering in bulk may have increased spending to avoid future stockouts.

3️⃣ Price Hikes Due to Weather Delays 🌦️
	•	Supply chain disruptions could have led to price increases, artificially inflating transaction values.
	•	Businesses might have ordered fewer items but at higher prices.

4️⃣ Shift in Customer Demographics 👥
	•	Maybe casual buyers left, but loyal buyers stayed.
	•	Are high-value customers the ones who continued purchasing
	
	
#### Forecasting overall purchase transaction trends associated with logistics efficiency
```{r}
monthly_data <- data %>%
  group_by(Month) %>%
  summarise(Avg_Transaction_Amount = mean(User_Transaction_Amount, na.rm=TRUE))
print(monthly_data)
```
```{r}
library(forecast)

ts_data <- ts(monthly_data$Avg_Transaction_Amount, start = c(2024,1), frequency = 12)

plot(ts_data, main="Monthly avg user transactions", ylab="Avg Transaction Amount", xlab="Month")
```
```{r}
fdata <- data %>%
  mutate(Date = as.Date(Timestamp),
         Month = format(Date, "%Y-%m"),
         Week = format(Date, "%Y-%U"))
print(fdata)
```
Since transaction behavior may depend on external factors, we should use:
	•	Shipment & Logistics Delays: Logistics_Delay, Waiting_Time, Logistics_Delay_Reason
	•	Environmental Factors: Temperature, Humidity, Traffic_Status
	•	Asset Performance: Asset_Utilization, Inventory_Level
	•	Customer Behavior: User_Purchase_Frequency
```{r}
regressor_vars <- c("Logistics_Delay", "Waiting_Time", "Temperature", "Humidity", "Traffic_Status","User_Purchase_Frequency","Asset_Utilization","Inventory_Level")
print(regressor_vars)
```
```{r}
library(dplyr)
library(lubridate)


monthly_data <- monthly_data %>%
  mutate(Month = as.Date(paste0("2024-", Month, "-01"), format="%Y-%b-%d"))

# Verify conversion
str(monthly_data$Month)
```


Use Time Series Regression with External Regressors



```{r}

# Convert timestamp and extract Year-Week
data <- data %>%
  mutate(Date = as.Date(Timestamp),  
         Week = format(Date, "%Y-%W"))  # Use "%W" for Monday-based weeks

# Ensure Week is treated as a proper date (assigning the start of the week)
data <- data %>%
  mutate(Week = as.Date(cut(Date, breaks = "week", start.on.monday = TRUE)))

# Aggregate data weekly
weekly_data <- data %>%
  group_by(Week) %>%
  summarise(
    User_Transaction_Amount = mean(User_Transaction_Amount, na.rm = TRUE),
    Logistics_Delay = mean(Logistics_Delay, na.rm = TRUE),
    Waiting_Time = mean(Waiting_Time, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    Humidity = mean(Humidity, na.rm = TRUE),
    User_Purchase_Frequency = mean(User_Purchase_Frequency, na.rm = TRUE),
    Asset_Utilization = mean(Asset_Utilization, na.rm = TRUE),
    Inventory_Level = mean(Inventory_Level, na.rm = TRUE),
    .groups = "drop"
  )

# Check results
head(weekly_data)
```

```{r}
x_data <- fread("smart_logistics_dataset.csv")
head(x_data)
```
  •	lag_1: Last week’s transaction amount
	•	lag_2: The amount from two weeks ago
	•	lag_3: The amount from three weeks ago

	•	The model uses these past values to predict future transactions.
```{r}
# Create Lag Features (Using past values to predict future)
x_data[, lag_1 := shift(User_Transaction_Amount, 1)]
x_data[, lag_2 := shift(User_Transaction_Amount, 2)]
x_data[, lag_3 := shift(User_Transaction_Amount, 3)]

```

```{r}
#Rolling statistics


# Compute rolling mean and standard deviation
x_data[, roll_mean_3 := rollapply(User_Transaction_Amount, width = 3, FUN = mean, align = "right", fill = NA)]
x_data[, roll_std_3 := rollapply(User_Transaction_Amount, width = 3, FUN = sd, align = "right", fill = NA)]
```

```{r}
x_data <- na.omit(x_data)
```

```{r}
#Train and Test
train_size <- floor(0.8 * nrow(x_data))
train <- x_data[1:train_size, ]
test <- x_data[(train_size+1): nrow(x_data), ]
```

```{r}

# Convert to Matrix for XGBoost
feature_cols <- c("lag_1", "lag_2", "lag_3", "roll_mean_3", "roll_std_3", 
                  "Logistics_Delay", "Waiting_Time", "Temperature", 
                  "Humidity", "User_Purchase_Frequency", 
                  "Asset_Utilization", "Inventory_Level")

train_matrix <- as.matrix(train[, feature_cols, with = FALSE])
test_matrix <- as.matrix(test[, feature_cols, with = FALSE])

train_labels <- train$User_Transaction_Amount
test_labels <- test$User_Transaction_Amount

# Train XGBoost Model
xgb_model <- xgboost(data = train_matrix, 
                     label = train_labels, 
                     nrounds = 100, 
                     objective = "reg:squarederror")

# Predict Future Values
predictions <- predict(xgb_model, test_matrix)

# Evaluate Performance
library(Metrics)
rmse_val <- rmse(test_labels, predictions)
mape_val <- mape(test_labels, predictions) * 100

print(paste("RMSE:", rmse_val))
print(paste("MAPE:", mape_val))
```

```{r}
future_weeks <- 16 

# Define last available date
last_week_date <- as.Date(max(weekly_data$Week))

# Generate future week dates
forecast_weeks <- seq(last_week_date + 7, by = "week", length.out = future_weeks)

# Prepare Future Regressors
future_regressors <- tail(x_data, future_weeks)
future_regressors <- as.matrix(future_regressors[, feature_cols, with = FALSE])

# Generate Forecasts
future_predictions <- predict(xgb_model, newdata = future_regressors)

# Create Forecast DataFrame
forecast_df <- data.frame(
  Week = forecast_weeks,
  Forecasted_Amount = future_predictions
)

# Convert 'Week' columns to Date format
forecast_df$Week <- as.Date(forecast_df$Week)

# Plot Forecasts
ggplot() +
  geom_line(data = weekly_data, aes(x = Week, y = User_Transaction_Amount), color = "blue") +  # Actual data
  geom_line(data = forecast_df, aes(x = Week, y = Forecasted_Amount), color = "red", linetype = "dashed") +  # Forecast
  labs(title = "XGBoost Forecast for User Transaction Amount",
       x = "Week", y = "Transaction Amount") +
  theme_minimal()
```
### Conclusion


The analysis of logistics data has revealed key trends and patterns affecting delivery performance and user transaction behavior. The most significant findings include:
	1.	Truck Performance: Trucks 10 and 3 experienced the highest number of delays, indicating potential inefficiencies or operational bottlenecks.
	2.	Seasonal Delays: Delays peaked in July, particularly due to weather-related disruptions, highlighting the impact of seasonal variations on logistics performance.
	3.	User Behavior Shift: Despite a sharp decline in user purchase frequency between August and September, the overall user transaction amount remained stable, suggesting that fewer but higher-value transactions were occurring.
	4.	Asset Utilization: Truck 5 had the lowest asset utilization at 76%, which may indicate under use or inefficiency in fleet management.
	5.	Overall Growth: A general increasing trend in logistics operations was observed from January to December, suggesting business expansion or improved operational efficiency.
	6.	Geographical Impact: The least amount of delays was observed in specific latitude-longitude ranges (0,100) to (50,100), potentially due to optimized routes or favorable geographical conditions.
	7.	Forecasting Insights: The predicted user transaction amount aligns well with the overall increasing trend and external factors, confirming the model’s reliability in capturing real-world logistics patterns.

These insights can be leveraged to optimize fleet performance, anticipate seasonal impacts, improve asset utilization, and refine forecasting models for better decision-making in logistics operations.
